pRolocGUI: An interactive tool for visualisation of organelle proteomics data
===========================================================

### Thomas Naake and Laurent Gatto*

#### *Computational Proteomics Unit, University of Cambridge


#### Foreword

This vignette describes the implemented functionality in the **pRolocGUI** 
package. The package is based on the MSnSet class defininitions of 
**MSnbase** [[1](#id1)] and on the functions defined in the package
**pRoloc** [[2](#id2)]
and is especially meant for the analyses of LOPIT [[3](#id3)] or PCP [[4](#id4)]
experiments.
To achieve reactivity and interactivity the function of pRolocGUI was written in 
[**shiny**](http://www.rstudio.com/shiny/) notation.
The implemented application facilitates a higher degree of 
interactivity with the underlying data, though, it is limited by its 
deterministic nature and will not offer the high flexibility given by the
use of a command based approach. Especially, pRolocGUI addresses to R
novices who have constraints of exploring their data by using the console.

pRolocGUI is under active development; current funtionality is evolving and 
new features will be added. This software is free and open-source. You are
invited to contact Thomas Naake in case you have any questions, suggestions 
or have found any bugs or typos. To reach a broader audience for more general
questions about proteomics analyses using R consider of writing to the 
Bioconductor list. 

-------------------
**Table of Contents**


1. [Introduction](#introduction)  
2. [Tabs of pRolocVis](#tabspRolocVis)    
  2.1. [Data](#tabspRolocVisData)  
  2.2. [PCA](#tabspRolocVisPCA)  
  2.3. [protein profiles](#tabspRolocVisPP)    
  2.4. [quantitation](#tabspRolocVisExprs)    
  2.5. [feature meta-data](#tabspRolocVisfData)   
  2.6. [sample meta-data](#tabspRolocVispData)     
  2.7. [search](#tabspRolocVisSearch)  
3. [*Display selection* widget](#display)    
  3.1. [PCA](#displayPCA)   
  3.2. [protein profiles](#displayPP)    
  3.3. [saved searches](#displaySaSe)    
  3.4. [query feature meta-data](#displayQuery)    
4. [References](#references)   
 


-------------------
<a id="introduction"></a> 
### 1. Introduction

Currently, you can start a pRolocVis application. 

Starting a pRolocVis application is easy as it gets. Just type 
```{r, eval = FALSE}
pRolocVis()
```
in your console window and press [Enter]. This will open a new tab in your 
Internet browser and after a bit you will see the application. You can also
pass directly a MSnSet (e.g. dunkley 2006) to pRolocVis by entering
```{r, eval = FALSE}
pRolocVis(object = dunkley2006)
```
To stop the application from running press [Esc] in the 
console (or use the "STOP" button when using RStudio) and close the browser 
tab, where pRolocVis is running. 

-------------------
<a id="tabspRolocVis"></a>
### 2. Tabs of pRolocVis
To optimize ease of use the interface of pRolocVis is subdivided in seven tabs:
* Data,  
* PCA,  
* protein profiles,     
* quantitation,   
* feature meta-data,   
* sample meta-data and  
* search.   

You browse through the tabs by simply clicking on them. Each tab
selected will have a different kind of appearance while some (PCA, protein 
profiles, quantitation and feature meta-data) share a common feature in the 
sidebar, the *Display selection* widget (see 3. *Display selection* widget for further details). 

<a id="tabspRolocVisData"></a>
#### 2.1. Data
The tab **_Data_** is the first tab you will see when you start pRolocVis. 
The appearance will differ depending on if you pass directly a MSnSet to 
pRolocVis. 

If you decide to start pRolocVis without assigning a MSnSet to the argument 
object the tab **_data_** offers possibilities to switch to an other MSnSet.
Currently you may choose between three example MSnSets (derived from experiments 
of Christoforou 2011, Dunkley et al. 2006, Tan et al. 2009) provided by the 
package **pRolocdata**  or to use an MSnSet of your choice (own data). 
Clicking on **_Browse..._** will open a dialog window with which you can select
your MSnSet and load it to pRolovVis. 
pRolocVis will check if the selected data file is of class MSnSet 
and if it is an .Rdata or .rda file and will denote a notification if there 
are any conflicts. 

If you pass an object to pRolocVis the tab data will show if the MSnSet was 
accepted by pRolocVis. 

<a id="tabspRolocVisPCA"></a>
#### 2.2. PCA
The tab **_PCA_** is characterized by its main panel which shows a PCA plot for 
the selected MSnSet and its sidebar panel which is compartimented into *Display selection*
and *Plot*. 

Let's forget about the compartiment *Display selection* here (see 3. 
*Display selection* widget for further details) and let's turn towards the *Plot* 
compartiment. Here we are able to adjust the appearance of the PCA plot in the
main panel. We are able to colour features (proteins) in matters of common
properties by changing the drop-down list **_colour_**, e.g. if we select 
for the
MSnSet originating from Christoforou 2011 the colour "markers", the features in 
the PCA plot will be coloured according to their organelle affiliation. 
As soon as we select another colour than "none"
two (or three) new items will be added to the *Plot* widget:   

(1) an item to change the symbol type, **_symbol type_**,   
(2) an item to manipulate legend properties, **_legend_** and 
**_position of legend_** , and  
(3) according to the MSnSet used an item to change the point size, 
**_point size_**. 

By selecting one of 
the variable names in the drop-down list of **_symbol type_** we will change
the 
symbol type of the features in the plot. By clicking on the check box to the 
left of **_legend_** a legend is added to the plot and by choosing one of the 
items in the drop-down list **_position of legend_** below we will change its 
position. The third item which might appear is the drop-down list 
**_point size_** to change the point size. pRolocVis will identify numeric 
values in the feature variable labels and will list them in the drop-down 
menu (this might be of interest if you have scores which give information 
about the quality of a phenoDisco analysis). 

By changing the drop-down lists of the items 
**_number of 1st principal component_** and 
**_number of 2nd principal component_** the x-values and 
y-values, respectively, will be rendered according to the new principal
components.

To zoom in and out change the drag and drop the little arrows of the slider
of the items **_zoom x-axis_** and **_zoom y-axis_**. This may be of great 
help when you want to identify points in dense clusters.

By clicking on **_Download Plot_** in the main panel below the PCA plot will
open
a dialog window with an interface on showing or saving the PCA plot as it is
displayed in the main panel.

<a id="tabspRolocVisPP"></a>
#### 2.3. protein profiles (has to be revised)
The tab **_protein profiles_** shows the protein profiles in the main panel 
(with an option of exporting the plot as it is shown in the main panel by 
clicking on the button **_Download Plot_**) and the *Display selection* widget and the
*Plot* widget in the sidebar panel. Have a look on section 
3. *Display selection* widget if you want to
retrieve information about how to use the *Display selection* widget. 

The *Plot* widget helps to manipulate the plots shown in the main panel.
Let's assume we want to have a look upon the protein profiles for the
proteins from which we know that they belong to the organelles endoplasmic
reticulum, the golgi apparatus, mitochondrion and the plasma membrane for 
the MSnSet originating from the experiments of Christoforou (2011). 
We have four organelles to look at, so we select "4" in the drop-down 
list **_number of plots to display_**. We will select
"markers" in the drop-down list and select "ER" (coding for endoplasmic 
reticulum) in the next drow-down list and leave the other two drop-down
lists as they are. To plot the next plot we have to change the slider 
**_Select number of plot to change_** to 
position 2 and for the shiny reactive expressions didn't receive yet a new
input we have to change the drop-down lists in order a new plot will be 
displayed, accordingly to our problem we will change the second drop-down list
to "Golgi" (coding for golgi apparatus). We proceed with the two remaining 
organelles as described before by changing firstly the slider to the next position
and by changing the drop-down lists addordingly to the organelles we want to
display.

We can change the two lower drop-down lists if we want to see some features 
compared to all features which are assigned to a variable name. To see display
results of phenoDisco we can select for the MSnSet from Christoforou 2011
"pd.markers" in the first drop-down list and "Golgi" as the organelle to look at. 
As the comparison we will choose in the third drop-down list "markers" and 
"Golgi" which will display the features in a transparent manner. 

<a id="tabspRolocVisExprs"></a>
#### 2.4. quantitation   
The tab **_quantitation_** displays the quantitation data for the proteins as a 
data table. 

In the main panel you can change the number of proteins shown per page and 
search both for proteins (or for the quantitation data). Also, you may sort 
the proteins by name or the quantitation data by clicking on the arrows 
on the top of the data table.

In the sidebar panel the *Display selection* widget is placed as well as radio buttons
to display all data or just selected features (see 3. *Display selection* widget for 
further details)

<a id="tabspRolocVisfData"></a>
#### 2.5. feature meta-data  
The tab **_feature meta-data_** displays the feature meta-data for the proteins
as a data table.

The layout of the tab is similar to the **_quantitation_** tab
and allows for sorting and searching in the feature meta-data of the MSnSet 
selected. 

The sidebar conprises the *Display selection* widget and radio buttons to 
show all or only selected features (see 3. *Display selection* widget for 
further details).

<a id="tabspRolocVispData"></a>
#### 2.6. sample meta-data    
The tab **_sample meta-data_** displays the sample meta-data for the 
experiment, the name of the isotopes used for tagging and the associated
fractions. 

<a id="tabspRolocVisSearch"></a>
#### 2.7. search  
pRolocVis allows to use past search results to display it in the PCA plot,
protein profiles and in the tabs **_quantitation_** and **_feature meta-data_**
(see 3. *Display selection* widget for further details if this is your intention). 
This ability requires an object **pRolocGUI_SearchResults** in the global 
environment which is of class "FeaturesOfInterest" or "FoICollection" 
(enter ?FeaturesOfInterest in the console for further details).

In case this objects exists it will automatically be loaded to pRolocVis and
its content is displayed in the tab **_search_**. Use the drop-down list in the
sidebar panel to browse through the different features of interest in case 
the object is of class "FoICollection". 

If no object called **pRolocGUI_SearchResults** exists in the global 
environment
you can assign an object to the global environment which is derived from the
experiment of Christoforou (2011) but will have no features. To do this,
click on **_Initialize saved searches_**, which will create a new 
FoICollection.

To save features of interest to the object in the global environment you need
to select features and add these to the FoICollection by entering an 
appropriate description in the text field (on the sidebar panel, which will
be useful to trace back to the underlying features and doesn't yet exist in 
the FoICollection) and by clicking on **_Create new features of interest_**.
This will add the selected features to the object **pRolocGUI_SearchResults**
in the global environment.

-------------------
<a id="display"></a>
### 3. *Display selection* widget
The *Display selection* widget is probably the most important implementation 
in pRolocVis for it enables the user to search for features in the MSnSet: 
you can do this by selecting points in the PCA plot, clicking on features 
in the tab **_protein profiles_**, using past searches and/or querying for 
features in the MSnSet data.

As you may have already seen there are four check boxes in the 
*Display selection* widget
which represent the before mentioned ways of searching features in the MSnSet. 
To activate the search for one specific method click on the check box left of
its description. It is also possible to select more than one at a time which 
allows for greater flexibility with regard to information retrieval.

<a id="displayPCA"></a>
#### 3.1. PCA 
If you decide to identify proteins in the PCA plot, change to the tab 
**_PCA_** and start clicking on features in the PCA plot 
(tipp: the zoom function may be of great expedient). The check box 
will be checked when you start clicking in the PCA plot. As soon as you have 
clicked on a feature it will be marked with a black circle around it. 
If you have selected a feature by accident or want to deselect it, 
just click again on the feature and it will be deselected. 

There are two possibilities to deselect all selected features: If you decide 
to remove all your features click on **_Clear features_**. Please keep in mind
that this step once carried out is irreversible. 
Besides that you are also able to simply blind out the selected features by 
deselecting the check box left of PCA in the *Display selection* widget. 
Internally, the  features are still stored, i.e. by clicking again on 
the check box you will see the selections again. Clicking on new proteins in 
the PCA plot will not check the check box again, so you have to do this
manually. The features selected are shared between the 
different tabs. Click on the tabs **_quantitation_** and 
**_feature meta-data_**
to have a look upon information about the selected features. For the case 
where you see all features in the data table change the radio buttons settings 
from **_all_** to **_selected_** at the lowermost widget in the sidebar. Here 
again, you can compose the features from different sources (PCA, protein 
profiles, saved searches and the text-based query search). 

If you display protein profiles in the tab **_protein profiles_** selected 
features will be displayed by black lines with a greater line width
(depending on the features shown). 

<a id="displayPP"></a>
#### 3.2. protein profiles
In principle the search for features in protein profiles is in accordance
with the search in the PCA plot. Though, bear in mind that you are only 
able to select features when "1" is selected in the drop-down list 
**_number of plots to display_**. Clicking on (or near) the points in the 
plot will select, clicking another time will deselect features. 
The features will only be shown when the check box left of 
**_protein profiles_** is activated. 

<a id="displaySaSe"></a>
#### 3.3. saved searches
Clicking on the check box to the left of **_saved searches_** will load the 
selected features of the class "FeaturesOfInterest". These will be displayed
in the PCA plot, in the plots for protein profiles (depending on the
displayed features) and will be available in the tabs **_quantitation_**
and **_feature meta-data_** for information retrieval. 
If there is an "FoICollection" loaded change the selected search 
result in the drop-down list **_Search result_** in the tab **_search_**; 
thus accordingly altering the selected features in the *Display selection* widget 
context. 

In pRolocVis there is no functionality implemented to remove features
from the object **pRolocGUI_SearchResults** in the global environment. The 
authors decided that it is not the task of a GUI to fulfill the requirements 
of this kind of data manipulation in a GUI, hence, the execution of removing
features of interests belongs to the field of the users responsibility.

<a id="displayQuery"></a>
#### 3.4. query feature-meta data
The *Display selection* widget offers the opportunity to query the feature meta-data
of the MSnSet for levels. The drop-down list consists of the item "protein",
which is by definition the rowname of the feature-meta data and depending on 
the data accession number, protein ID, protein description, assigned markers
(varying on the underlying MSnSet). 

Let's assume we want to look in the MSnSet which was derived from experiments 
of Christoforou (2011) for all proteins which are assigned by experimental
evidence to the organelle "plasma membrane". We ensure ourselves that 
"Christoforou 2011" is selected in the tab **_Data_** and change to a tab where
the *Display selection* widget is loaded. We activate the check box to the left of 
**_level_** and select "marker" in the upper drop-down list (for we are 
looking for
the organelle). In the next drop-down list below we select "PM" which codes
for "plasma membrane". Next, we click on **_Submit selected level_**, which
will
highlight all features which are assigned to "PM" for the variable name
"marker". To remove the selected features from the internal assignment we 
have to reset the search by clicking on **_Clear features_**. Of course,
we can also add other features: If we want to add all features which are 
assigned to the Golgi apparatus we simply select "Golgi" in the lower drop-down
list and click on **_Submit selected level_** to save internally the selected 
features. 

It is relatively easy to find levels when the drop-down list for these levels
is short (as it is for "markers" in the MSnSet originating from the 
experiments of Christoforou (2011)). But how should we proceed when we want to
look for a special protein, e.g. OCAD1? The drop-down list for the variable 
name "protein" is very long and it is time consuming to scroll through the 
whole list and look for our protein of interest. Therefore, we can just enter 
OCAD1 in the text input field **_Search for_** in between the two drop-down 
lists and we will get the protein of interest (we are also able to query for
protein names which have the string "OC" in their name which will limit the 
drop-down list to all proteins which have this specific string). By clicking on 
**_Submit selected level_** we save internally the selected feature(s).

<a id="references"></a>
#### 4. References
<a id="id1"></a>
[1] Laurent Gatto and Kathryn S. Lilley. MSnbase - an R/Bioconductor
package for isobaric tagged mass spectrometry data visualization,
processing and quantitation. Bioinformatics 28, 288-289 (2011).

<a id="id2"></a>
[2] Gatto L, Breckels LM, Wieczorek S, Burger T, Lilley KS. Mass-spectrometry-
based spatial proteomics data analysis using pRoloc and pRolocdata. Bioin-
formatics. 2014 Feb 5.

<a id="id3"></a> 
[3] Tom P. J. Dunkley, Svenja Hester, Ian P. Shadforth, John Runions,
Thilo Weimar, Sally L. Hanton, Julian L. Griffin, Conrad Bessant, Federica
Brandizzi, Chris Hawes, Rod B. Watson, Paul Dupree, and Kathryn S. Lilley.
Mapping the arabidopsis organelle proteome. Proc Natl Acad Sci USA, 
103(17):6518–6523, Apr 2006. doi: 10.1073/pnas.0506958103. 
URL http://dx.doi.org/10.1073/pnas.0506958103.

<a id="id4"></a> 
[4] Leonard J. Foster, Carmen L. de Hoog, Yanling Zhang, Yong Zhang, Xiaohui Xie,
Vamsi K. Mootha, and Matthias Mann. A mammalian organelle map by protein
correlation profiling. Cell, 125(1):187–199, Apr 2006. 
doi: 10.1016/j.cell.2006.03.
022. URL http://dx.doi.org/10.1016/j.cell.2006.03.022.




